---
title: "Nuclear Explosions"
description: |
  This a small story about nuclear detonations told with some exploratory data visualization and analysis.
author:
  - name: Niketh Gamage
    url: {}
date: 04-04-2021
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE, error=TRUE, message=FALSE, warning=FALSE)
```

```{r libraries,include=FALSE}
library(tidyverse)     # for data cleaning and plotting
library(gardenR)       # for Lisa's garden data
library(lubridate)     # for date manipulation
library(openintro)     # for the abbr2state() function
library(palmerpenguins)# for Palmer penguin data
library(maps)          # for map data
library(ggmap)         # for mapping points on maps
library(gplots)        # for col2hex() function
library(RColorBrewer)  # for color palettes
library(sf)            # for working with spatial data
library(leaflet)       # for highly customizable mapping
library(ggthemes)      # for more themes (including theme_map())
library(plotly)        # for the ggplotly() - basic interactivity
library(gganimate)     # for adding animation layers to ggplots
library(gifski)        # for creating the gif (don't need to load this library every time,but need it installed)
library(transformr)    # for "tweening" (gganimate)
library(shiny)         # for creating interactive apps
library(patchwork)     # for nicely combining ggplot2 graphs  
library(gt)            # for creating nice tables
library(rvest)         # for scraping data
library(robotstxt)     # for checking if you can scrape data
theme_set(theme_minimal())
```

## Overview

```{r include=FALSE}
nuclear_explosions <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-08-20/nuclear_explosions.csv")
data(nuclear_explosions)
# levels(factor(nuclear_explosions$country))
```


Nuclear weapons are the deadliest weapons known to man. In this dataset([which can be found here](https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-08-20)), we will be exploring the history of nuclear weapons detonation in the world. The data used for this analysis spans from 1945 - 1998, so the results of the analysis presented here are as of 1998.

I will go about answering these questions: 

1. What are the countries with nuclear weapons capability and when did they achieve this?

2. Which country has detonated the most nuclear weapons? 

3. How have these detonation patterns varied over time - what effect did the fall of the Soviet Union have  on this?

4. Where were these nuclear weapons detonated?

5. Which countries definitevely have a fusion bomb(2nd generation of nukes)?

6. How did methods of testing them change over time?

7. Simple application that lets you compare the countries nuclear detonation activity over time.




### 1. When did each country first test a nuclear weapon?
```{r echo = FALSE}
ggplotly(nuclear_explosions %>% 
  group_by(country) %>% 
  summarise(first_bomb = min(year)) %>% 
  ggplot()+
  geom_hline(yintercept=0, 
                color = "black", size=0.3)+ 
  geom_point(aes(x = first_bomb, y=0, color = country))+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line.y = element_blank(), axis.ticks.y = element_blank(), axis.text.y = element_blank())+ 
  labs(y = "",x="", title = "First Nuclear Detonation by Country") )

  
```

Here we see a timeline which shows all the nuclear powers in the world(as of 1998) and when they first reached this capability. In order of first detonation, these countries are the USA, USSR, UK, France, China, India and Pakistan. 


### 2. Which country has detonated the most nuclear weapons?

```{r echo=FALSE}

# nuclear_explosions %>% 
#   count(country)

country_count_plot<- nuclear_explosions %>% 
  count(country) %>% 
  ggplot(aes(x = n, y = fct_reorder(country,n), fill = country))+
  geom_col() +
  labs(y = "", x= "", title = "Total nuclear bomb detonations")+
  theme(panel.grid.major.y = element_blank())+
  theme(legend.position = "none")
 
ggplotly(country_count_plot )

```

The United States and the USSR comprise a large majority of the total nuclear detonations made. 





### 3. How have these detonation patterns varied over time - what effect did the fall of the Soviet Union have  on this?
```{r echo = FALSE}
# nuclear_explosions%>%
#   group_by(year) %>% 
#   count()

ggplotly(nuclear_explosions %>%
  group_by(year) %>% 
  count() %>% 
  ggplot(aes(x = year, y = n)) +
  geom_col(fill = "purple")+
  scale_x_continuous(breaks = c(1945,1950,1960,1970,1980,1990))+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  labs(title = "Total number of nuclear explosions by year"))


```

This is a very interesting distribution. We can see how the first weapons were developed in the mid 1940's after which the testing heavily accelerated in the 1950's before coming to an abrupt stop in 1959 (apparently due to a nuclear testing moratorium between the US, USSR and the UK from November 1958 to August 1961). However testing increased back up and remained at a consistently high level during the 2 decades from 1960 - 1980 which also corresponded to the peak of the cold war. 

1962 in particular was a crazy year with a total of 178 detonations performed - a nuclear explosion every other day. It is also interesting that this was the year where the world saw the Cuban Missile Crisis in which the tensions between the USA and the USSR peaked and the world appeared to be at the edge of a nuclear calamity. 

In the mid to late 1980's, testing dropped down significantly and plummeted furthermore in the 1990's until there were only a couple of tests each year. The Soviet Union dissolved between 1988-1991 and it is interesting to see how the nuclear testing in the world fell sharply during and after this period.

### 4. Where were these tested by each country? And how did this look over time?
```{r include=FALSE, warning=FALSE}
world <- get_stamenmap(
    bbox = c(left = -180, bottom = -57, right = 179, top = 82.1), 
    maptype = "terrain",
    zoom = 2)
```

```{r echo = FALSE, warning=FALSE}
(ggmap(world)+
  geom_point(data = nuclear_explosions, aes(x=longitude, y=latitude, alpha = 0.01, color = country)) +
  labs(title = "Geographical Distribution of Nuclear Bomb Detonations ", y = "", x=""))+
  guides(alpha = FALSE)+
  theme(axis.text = element_blank())

```

The US appears to have tested in the mainland(mostly in Nevada) and in the Pacific over water and over some Island and some tests in the South Athlantic. The only 2 bombs used in combat ever can be observed over Japan. USSR tested all over their mainland, concentrated in small areas in the south in what appears to be modern day Kazakhstan, Uzbekistan and other former Soviet terriories and in the Novaya Zemlaya islands in the north. UK never tested on mainland, instead tested in Australia, off the coast of Australia and interstingly in mainland America(Nevada). France never tested in their mainland, but instead in what appears to be Algeria and in the South Pacific. China, India and Pakistan appear to have tested in their respective mainlands exclusively.

Here we have a small animation displaying these detonations occuring over time around the world.


```{r echo = FALSE, warning=FALSE}
ggmap(world)+
  geom_point(data = nuclear_explosions, aes(x=longitude, y=latitude, alpha = 0.01, color = country , size = yield_upper))+
  transition_reveal(year)+
  labs(subtitle = "Year:{frame_along} ", title = "Nuclear explosions around the world over time")+
  exit_shrink()+
  theme(legend.position = "top",
        legend.title = element_blank(),
        )


```



### 5. Which countries definitevely have a fusion bomb(2nd generation of nukes) and what are the yields of the most destructive bombs for each country?

Fusion bombs are the 2nd generation of nuclear bombs - they use a fission bomb to activate a fusion reaction which releases a vast amount of energy. The destructive capabilities of fusion bombs can be several order of magnitudes higher than a regular fission bomb.
Anything above 500kT yield is definitely a thermonuclear bomb(fusion bomb as opposed to a fission bomb) so here I explored which countries definitively have this capability (as of 1998).


```{r echo=FALSE}
nuclear_explosions %>% 
  filter(yield_upper > 500) %>% 
  count(country) %>% 
  ggplot()+
  geom_col(aes(y=fct_reorder(country,n), x = n), fill = "red") + 
  labs(x = "amount of bombs detonated with yield > 500kT TNT", y="")+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

```

We can observe at least 100 large fusion bomb tests in total. These have primarily undertaken by the USA and USSR but China, UK and France have definitely shown capability too (as of 1998) .

Then we explore the yield of the strongest bombs produced by each country

```{r echo = FALSE}
ggplotly( nuclear_explosions %>% 
  group_by(country) %>% 
  slice_max(order_by = yield_upper,n= 1) %>% 
  select(country, yield_upper) %>% 
  unique() %>% 
  ggplot(aes(x = fct_reorder(country,yield_upper), y = yield_upper, color= country))+
  geom_point()+
  labs(y = " Max Yield (kT of TNT)", x = "", title = "Most Powerful Bombs by Country")+
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), legend.position = "none"))
```



### 6. How did methods of testing them change over time?

We focus on the 5 most common methods of testing out of a total 20. These 5 were: AIRDROP, ATMOSPH, SHAFT, SHAFT/GR, TUNNEL. AIRDROP is a device dropped from an aircraft and exploded in the atmosphere and ATMOSPH is also a method of testing performed above ground/water. SHAFT, SHAFT/GR, and TUNNEL are all underground methods of testing with their own different procedures.The results in how common they were over time are as follows.


```{r echo = FALSE}
# nlevels(factor(nuclear_explosions$type))
common_type <- nuclear_explosions %>% 
  count(type) %>% 
  slice_max(order_by = n, n=5) %>% 
  select(type) %>% 
  pull(type)

ggplotly(nuclear_explosions %>% 
  filter(type %in% common_type) %>% 
  group_by(year) %>% 
  count(type) %>% 
  ggplot(aes(x = year, y = n, color = type))+
  geom_line() +
  scale_color_viridis_d()+
  theme(panel.grid = element_blank(), legend.title = element_blank())+
  labs(title = "Most common nuclear bomb testing methods over time"))

```

It is interesting to see that in the earlier days of testing ATMOSPH and AIRDROP - which are the two above ground/water methods of testing were the most popular. However, in the 1960's and 1970's these methods of testing plummeted and methods of testing that were underground became more common (sidenote: this was apparently to minimize any health hazards from radioactive dust and waste being blown far by the wind into populated areas.)

### 7. Very simple application that lets you compare each countries nuclear detonation activity over time.

[Link to app](https://niks27.shinyapps.io/Nuclear_Explosions_app/). 

